---
title: "Example R notebook"
output:
  pdf_document: default
  html_notebook: default
---

# Loading packages 

The very first thing that you would like to do is loading the required R libraries.
Lets check whether you have successfully installed all the packages.

```{r}
library(tidyverse)
library(emmeans)
library(brms)
library(tidybayes)
library(bayestestR)
```

You will probably see some start message, and several warnings. If you don't see any errors, then with 99% you are probably fine with this step.

# Obtaining the dataset.

Most of the datasets will be availabe through the course online repository (or some other repositories). 
Lets try to get the example dataset.

```{r}
library(foreign)
crime_data <- read.dta("https://stats.idre.ucla.edu/stat/data/crime.dta")
```

# Initial lookup at the data

There are several ways to look at the dataset. Lets try a simple `glimpse`.

```{r}
crime_data %>% 
  glimpse()
```

# Plotting some of the distributions.

We are interested in the crime level through states. Lets plot a simple histogram.

```{r}
crime_data %>% 
  ggplot(aes(crime)) +
  geom_histogram()
```

The distribution looks fine appart from a single outlier.
Lets plot one the predictors of crime - poverty.

```{r}
crime_data %>% 
  ggplot(aes(poverty)) +
  geom_histogram()
```

Now lets plot crime against poverty.

```{r}
crime_data %>% 
  ggplot(aes(poverty, crime)) +
  geom_point()
```


# Creating a simple Bayesian regression model.

Lets fit crime against poverty. This is your `Hello world!` to Bayesian modelling.
Don't worry if you don't understand what you are doing or what is happening. This just a quick intro to get you the feeling of how we will work through the course. 

```{r}
fit <- brm(crime ~ poverty,
           data = crime_data,
           prior = prior(normal(0, 10), class = b))
```


# Summarising our results

Lets summarise our results.

```{r}
fit
```

It seems like poverty is positively and reliably associated with crime level (at the state level).

# Plotting our posterior.

Lets see to what extent the obtained posterior distribution supports positive association.

```{r}
pd_fit <- p_direction(fit)
plot(pd_fit)
```

# Plotting posterior against prior

Now lets compare posterior distribution to prior distribution.

```{r}
bf_fit <- bayesfactor_parameters(fit)
plot(bf_fit)
```

# Plotting predictions against data

Now lets plot the fitted regression line against the collected data.

```{r}
marginal_effects(fit, effects = "poverty") %>% 
  plot(points = T)
```

If you would like to fit predicted crime level given the poverty level it is also quite simple.

```{r}
marginal_effects(fit, effects = "poverty", method = "predict") %>% 
  plot(points = T)
```

If you find the code in this notebook hard to follow, DON'T PANIC. It will get easier with time.
I would also recommend you to go through some online tutorials. For example Hadley Wickham's [R for Data Science](https://r4ds.had.co.nz/).

