---
title: "Final exam"
output: html_notebook
---

Run the code below to setup your R envirnoment to the exam.

```{r}
library(haven)
library(tidyverse)
library(brms)
library(emmeans)

df <- read_sav("roundPaper.sav")

df %>% 
  mutate(zsjus = (sjus - mean(sjus, na.rm=T))/sd(sjus, na.rm=T),
         zcm = (cm - mean(cm, na.rm=T))/sd(cm, na.rm=T),
         zround = (round - mean(round, na.rm=T))/sd(round, na.rm=T)) -> df
df %>% 
  mutate(voting = factor(polit1C, levels=1:3, labels=c("right","centerleft","nonvoting"))) -> df
```

# Task 1

Construct a prior object

```{r}
prior <- prior()
```

Fit a model

```{r}
fit1 <- brm()
```

Print the model

```{r}
fit1
```

Conduct posterior predictive checks

i.
```{r}
pp_check()
```

ii.
```{r}
pp_check()
```

iii.
```{r}
pp_check()
```

iv.
```{r}
pp_check()
```

v.
```{r}
pp_check()
```

Fit a model with Student t family

```{r}
fit1_alt <- brm()
```

Print the alternative model

```{r}
fit1_alt
```

## Task 2

Construct a prior object or use the prior from previous task

```{r}
prior <- prior()
```

Fit a model with interaction

```{r}
fit2_int <- brm()
```

Fit a model without interaction

```{r}
fit2_non_int <- brm()
```

Compute WAIC for the model with interaction

```{r}
fit2_int <- add_criterion()
```

Print WAIC for the model with interaction

```{r}
fit2_int$waic
```

Compute WAIC for the model without interaction

```{r}
fit2_non_int <- add_criterion()
```

Print WAIC for the model without interaction

```{r}
fit2_non_int$waic
```

Compute the difference of WAICs

```{r}
loo_compare()
```

Plot the interaction

```{r}
marginal_effects()
```

Compute simple slopes for the interaction

```{r}
emtrends()
```

