---
title: "Class 2: Bayes theorem and Bayesian updating"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F, fig.width = 6, fig.height = 4)
```

## Bayes theorem

$$Pr(A | B) = \frac{Pr(B | A) \times Pr(A)}{Pr(B)}$$

## Bayes theorem

- Suppose there is a test for vampirism that correctly detects vampirism in 90% of cases, i.e. given that you test a vampire, in 95 out of 100 cases it will return positive result.
- The test has some flows, because it can also return positive result when given to a human. Luckilly, the chance of such a result is small, 20%, i.e. given that you test a human in 10 out of 100 cases it will return positive result.
- We also know that vampires are quire rare, there is only 1 in 100 individuals.
- What is the chance that someone you tested and who obtained positive test result is a vampire.

## Bayes theorem

- $Pr(positive | vampire) = .90$ and this implies $Pr(negative | vampire) = .10$
- $Pr(positive | human) = .10$ and this implies $Pr(negative | human) = .90$ 
- $Pr(vampire) = .01$, and this implies $Pr(human) = .99$
- $Pr(vampire | positive) = ...$

## Bayes theorem

$$Pr(vampire | positive) = \frac{Pr(positive | vampire) \times Pr(vampire)}{Pr(positive)}$$

$Pr(positive) = ???$

## Bayes theorem

- How frequently the test will return positive result?
- Suppose you give a test to 1000 individuals:
  - If $Pr(vampire) = .10$, then 10 will be vampires, and 990 will be humans.
  - If $Pr(positive | vampire) = .90$, then among 10 vampires, you obtain 9 positive results.
  - If $Pr(positive | human) = .10$, then among 990 humans, you obtain 99 positive resutls.
  - In sum you will obtain 108 positive results out of 1000 tests. Then $Pr(positive) = .108$

## Bayes theorem 

- In mathematical notation you can write the formula for faster computation.

$$Pr(p.) = [Pr(p. | v.) \times Pr(v.)] + [Pr(p. | h.) \times Pr(h.)]$$

## Bayes theorem

$$Pr(v. | p.) = \frac{Pr(p. | v.) \times Pr(v.)}{[Pr(p. | v.) \times Pr(v.)] + [Pr(p. | h.) \times Pr(h.)]}$$

## Bayes theorem

$Pr(vampire | positive) = \frac{.90 \times .01}{[.90 \times .01] + [.10 \times .99]}$

$Pr(vampire | positive) = \frac{.90 \times .01}{.108}$

$Pr(vampire | positive) = \frac{.90 \times .01}{.108}$

$Pr(vampire | positive) = .0833$ 

## Your turn



## Bayes theorem

$$Posterior = \frac{Likelihood \times Prior}{Average Likelihood}$$

## Bayesian updating

- Suppose you have build a perfect detector of vampires.
- You start examining randomly encountered people.
  - What is a random sample?
- You have obtained a sample of 10 independent records.
  - What is the definition of independent records?
- Vampire, Human, Human, Vampire, Human, Human, Human, Human, Human, Human

## Bayesian updating

```{r}
library(tidyverse)
ggplot(data.frame(p = seq(0,1, length.out = 100)), aes(x=p))+
  stat_function(fun = function(x) dbeta(x, 1,1), size=1.5, linetype=2)+
  stat_function(fun = function(x) dbeta(x, 2,1), size=1.5, linetype=1)+
  ggtitle("V")+
  labs(y="Plausibility",x="Proportion of vampires")+
  ggpubr::theme_pubr()+
  theme(axis.text.y = element_blank())
```

## Bayesian updating

```{r}
library(tidyverse)
ggplot(data.frame(p = seq(0,1, length.out = 100)), aes(x=p))+
  stat_function(fun = function(x) dbeta(x, 2,1), size=1.5, linetype=2)+
  stat_function(fun = function(x) dbeta(x, 2,2), size=1.5, linetype=1)+
  ggtitle("V H")+
  labs(y="Plausibility",x="Proportion of vampires")+
  ggpubr::theme_pubr()+
  theme(axis.text.y = element_blank())
```


## Bayesian updating

```{r}
library(tidyverse)
ggplot(data.frame(p = seq(0,1, length.out = 100)), aes(x=p))+
  stat_function(fun = function(x) dbeta(x, 2,2), size=1.5, linetype=2)+
  stat_function(fun = function(x) dbeta(x, 2,3), size=1.5, linetype=1)+
  ggtitle("V H H")+
  labs(y="Plausibility",x="Proportion of vampires")+
  ggpubr::theme_pubr()+
  theme(axis.text.y = element_blank())
```


## Bayesian updating

```{r}
library(tidyverse)
ggplot(data.frame(p = seq(0,1, length.out = 100)), aes(x=p))+
  stat_function(fun = function(x) dbeta(x, 2,3), size=1.5, linetype=2)+
  stat_function(fun = function(x) dbeta(x, 3,3), size=1.5, linetype=1)+
  ggtitle("V H H V")+
  labs(y="Plausibility",x="Proportion of vampires")+
  ggpubr::theme_pubr()+
  theme(axis.text.y = element_blank())
```

## Bayesian updating

```{r}
library(tidyverse)
ggplot(data.frame(p = seq(0,1, length.out = 100)), aes(x=p))+
  stat_function(fun = function(x) dbeta(x, 3,3), size=1.5, linetype=2)+
  stat_function(fun = function(x) dbeta(x, 3,4), size=1.5, linetype=1)+
  ggtitle("V H H V H")+
  labs(y="Plausibility",x="Proportion of vampires")+
  ggpubr::theme_pubr()+
  theme(axis.text.y = element_blank())
```

## Bayesian updating

```{r}
library(tidyverse)
ggplot(data.frame(p = seq(0,1, length.out = 100)), aes(x=p))+
  stat_function(fun = function(x) dbeta(x, 3,4), size=1.5, linetype=2)+
  stat_function(fun = function(x) dbeta(x, 3,5), size=1.5, linetype=1)+
  ggtitle("V H H V H H")+
  labs(y="Plausibility",x="Proportion of vampires")+
  ggpubr::theme_pubr()+
  theme(axis.text.y = element_blank())
```

## Bayesian updating

```{r}
library(tidyverse)
ggplot(data.frame(p = seq(0,1, length.out = 100)), aes(x=p))+
  stat_function(fun = function(x) dbeta(x, 3,5), size=1.5, linetype=2)+
  stat_function(fun = function(x) dbeta(x, 3,6), size=1.5, linetype=1)+
  ggtitle("V H H V H H H")+
  labs(y="Plausibility",x="Proportion of vampires")+
  ggpubr::theme_pubr()+
  theme(axis.text.y = element_blank())
```

## Bayesian updating

```{r}
library(tidyverse)
ggplot(data.frame(p = seq(0,1, length.out = 100)), aes(x=p))+
  stat_function(fun = function(x) dbeta(x, 3,6), size=1.5, linetype=2)+
  stat_function(fun = function(x) dbeta(x, 3,7), size=1.5, linetype=1)+
  ggtitle("V H H V H H H H")+
  labs(y="Plausibility",x="Proportion of vampires")+
  ggpubr::theme_pubr()+
  theme(axis.text.y = element_blank())
```

## Bayesian updating

```{r}
library(tidyverse)
ggplot(data.frame(p = seq(0,1, length.out = 100)), aes(x=p))+
  stat_function(fun = function(x) dbeta(x, 3,7), size=1.5, linetype=2)+
  stat_function(fun = function(x) dbeta(x, 3,8), size=1.5, linetype=1)+
  ggtitle("V H H V H H H H H")+
  labs(y="Plausibility",x="Proportion of vampires")+
  ggpubr::theme_pubr()+
  theme(axis.text.y = element_blank())
```

## Bayesian updating

```{r}
library(tidyverse)
ggplot(data.frame(p = seq(0,1, length.out = 100)), aes(x=p))+
  stat_function(fun = function(x) dbeta(x, 3,8), size=1.5, linetype=2)+
  stat_function(fun = function(x) dbeta(x, 3,9), size=1.5, linetype=1)+
  ggtitle("V H H V H H H H H H")+
  labs(y="Plausibility",x="Proportion of vampires")+
  ggpubr::theme_pubr()+
  theme(axis.text.y = element_blank())
```